Exploring probability distributions for bivariate temporal granularities

### 2019-11-27 main.Rmd
split time granularity sections

# Time granularities{#sec:timegrandef}

<!-- general calendar categorization -->
Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Classes of time abstractions may be based on arrangement or hierarchical order of time, discussed in further details in \autoref{sec:arrangement} and \autoref{sec:order}.

## Arrangement: Linear vs. Circular vs. aperiodic {#sec:arrangement}

The arrangement of the time domain can result in deconstructing time in linear, circular or aperiodic (nearly circular) ways. Time granularities are **linear** if they respect the linear progression of time. Examples include hours, days, weeks and months. However, periodicity is very common in all kinds of data. Time granularities which can accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Hence, time granularities can also be **circular** or **aperiodic** depending on if the mapping of the linear granularities is regular or irregular. Examples of circular time granularities include hour of the day, and day of the week, whereas examples for aperiodic time granularities can be day of the month or public holidays.

@@

In the following sections, we provide a formalism to all of these abstractions and identify their relationships and symbolic representations.

### Linear {#sec:linear-gran-def}

@@

#### Relationships between two linear granularities and association to Periodicity

@@

\begin{timedom}\label{def:finerthan}
A linear granularity G is finer than a  linear granularity H, denoted $G \preceq H$, if
for each index i, there exists an index j such that $G(i) \subset H(j)$.
\end{timedom}

@@

According to this definition, $G \trianglelefteq H,$ if each granule H (j) is the union of some
granules of G. For example, $Days \trianglelefteq Months$. However, this  relationship is not fully described until we associate it with periodicity. For example, $Days \trianglelefteq Months,$, since each month is a grouping of a number of days, however it is also a periodical grouping. If leap year is ignored, the periodicity of the grouping is 1 year, since each month would be defined as the grouping of the same number of days (31, 28, or 30, depending on the month) every year. Considering leap years and all their exceptions, the period becomes 400 years. 
\begin{timedom}\label{def:periodical}
A granularity H is periodical with respect to a granularity G if
(1) $G \trianglelefteq H,$, and
(2) there exist $R,P \epsilon Z+$, where R is less than the number of granules of H, such
that for all $i \epsilon Z$, if $H(i) = \bigcup_{j \in S}G(j)$ and $H (i + R) \neq \phi$ then
$H (i + R) = \bigcup_{j \in S} G(j + P)$.
\end{timedom}

@@

This formula can be explained observing that H (j ) is either one of H (0), . . . , H (R - 1) or, for the periodicity of H, is one of these granules “shifted” ahead or behind on the time line of a finite number of periods. If $H (j ) \epsilon {H (0), . . . , H (R - 1)}$, then the formula ‘j mod R’ defines the index (among those in {0,...,R - 1}) of the granule that must be shifted to obtain H (j).The number of periods each granule of G composing H (j mod R) should be shifted is given by $\lfloor j/R \rfloor$.

@@

@@

\begin{timedom}\label{def:BG}
**Definition: Bottom granularity** Given a granularity order relationship g-rel and a set of granularities having the same time domain, a granularity G in the set is a bottom granularity with respect to g-rel, if G g-rel H for each granularity H in the set. 
\end{timedom}

**Example:** Given the set of all granularities defined over the time domain (Z; <), and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

@@

#### Computation of linear time granularities

@@

- **The grouping operation** : Let G be a full-integer labeled granularity, and m a positive integer. The grouping operation $Group_m(G)$ generates a new granularity G, by partitioning the granules of G into m-granule groups and making each group a granule of the resulting granularity.
More precisely, $G = Group_m(G)$ is the full-integer labeled granularity such that for
each integer i, $G(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G(j)$.

@@

- **The altering-tick operation** : Let G1, G2 be full-integer labeled granularities, and l, k, m integers, where G2 partitions G1, and $1 \leq l \leq m$. The altering-tick operation $Alter^{m}_{l,k} (G2, G1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the lth granule of each group has |k| additional (or fewer when k < 0) granules of G2.

@@

### Cyclic {#sec:cyclic-gran-def}

@@

#### Circular {#sec:circular-gran-def}

@@

\begin{equation} \label{eq:eq2}
C_{BG, G}(z) = BG(z\mod k(BG,G)) \forall z \in \mathbb{Z}^+,
BG(z) \neq \phi
\end{equation}
where,  
z denotes the index set.  
BG denotes the index of the tsibble (bottom granularity).  
G be a linear granularity, periodic with respect to BG.    
k(BG,G) be the period length of the bottom granularity with respect to linear granularity G.  
\end{timedom}

@@

### Aperiodic {#sec:aperiodic-gran-def}

@@

\begin{equation}\label{eq:eq3}
A_{BG, AG}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert), z \in T_k
\end{equation}
where,  
z denotes the index set (index of BG),
R' denotes the number of granules of AG in each period of BG,  
$T_w$ are the sets of indices of BG describing $AG(w)$ such that $AG(w) = \bigcup_{z \in T_w}BG(z)$, and $T_w = 0$ for $w \epsilon Z_{<0}$  
$\vert T_w \vert$ is the cardinality of set $T_w$.

@@

@@

## Order: Single vs. Multiple {#sec:order}

@@

## Harmony and Clashes

@@

@@

Most commonly used techniques to display a distribution of data include the histogram (Karl Pearson), which shows the prevalence of values grouped into bins and the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg, 1978] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. Moreover, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg, 2006] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail. “outliers” in letter value plots are those observations beyond the most extreme letter value. The letter values are shown till the depths where the letter values are reliable estimates of their corresponding quantiles and hence might lead to a lot of letter values being shown and leading to overload of information in one plot.

@@

Traditional ways to visualize densities include violin plots [@Hintze1998-zi, 1998]. The shape of the violin represents the density estimate of the variable. The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories, enabling easier detection of clusters or bumps within a distribution.

@@

Even when there are no rarely occuring events, number of observations might vary hugely across within or across each panel. This might happen due to missing observations in the data or uneven locations of events in time domain. In such cases, the visualizations that use kernel density estimates should be used with caution as sample size would directly affect both the variance of the estimator and the confidence intervals. Measures such as Gini's coefficient may be used to see if sample sizes vary widely across panels.

>>>>

# Linear Time granularities {#sec:lineartime}

<!-- general calendar categorization -->
<!-- Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Classes of time abstractions may be based on arrangement or hierarchical order of time, discussed in further details in \autoref{sec:arrangement} and \autoref{sec:order}. -->

Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Time granularities are **linear** if they respect the linear progression of time. Examples include hours, days, weeks and months.

## Definitions and Relationships

<!-- Arrangement: Linear vs. Circular vs. aperiodic {#sec:arrangement} -->

<!-- The arrangement of the time domain can result in deconstructing time in linear, circular or aperiodic (nearly circular) ways.  -->

@@

<!-- In the following sections, we provide a formalism to all of these abstractions and identify their relationships and symbolic representations.  -->

<!-- ### Linear {#sec:linear-gran-def} -->

@@

@@

\begin{timedom}\label{def:finerthan}
A linear granularity G is finer than a  linear granularity H, denoted $G \preceq H$, if
for each index i, there exists an index j such that $G(i) \subset H(j)$.
\end{timedom}.

**Dicussion:**

@@

**Dicussion:** According to this definition, $G \trianglelefteq H,$ if each granule H (j) is the union of some
granules of G. For example, $Days \trianglelefteq Months$. However, this  relationship is not fully described until we associate it with periodicity. For example, $Days \trianglelefteq Months,$, since each month is a grouping of a number of days, however it is also a periodical grouping. If leap year is ignored, the periodicity of the grouping is 1 year, since each month would be defined as the grouping of the same number of days (31, 28, or 30, depending on the month) every year. Considering leap years and all their exceptions, the period becomes 400 years.

\begin{timedom}\label{def:periodical}
A granularity H is periodical with respect to a granularity G if
(1) $G \trianglelefteq H,$, and
(2) there exist $R,P \epsilon Z+$, where R is less than the number of granules of H, such
that for all $i \epsilon Z$, if $H(i) = \bigcup_{j \in S}G(j)$ and $H (i + R) \neq \phi$ then
$H (i + R) = \bigcup_{j \in S} G(j + P)$.
\end{timedom}

@@

**Dicussion:** This formula can be explained observing that H (j ) is either one of H (0), . . . , H (R - 1) or, for the periodicity of H, is one of these granules “shifted” ahead or behind on the time line of a finite number of periods. If $H (j ) \epsilon {H (0), . . . , H (R - 1)}$, then the formula ‘j mod R’ defines the index (among those in {0,...,R - 1}) of the granule that must be shifted to obtain H (j).The number of periods each granule of G composing H (j mod R) should be shifted is given by $\lfloor j/R \rfloor$.

@@

**Dicussion:**

@@

\begin{timedom}\label{def:BG}
Bottom granularity - Given a granularity order relationship g-rel and a set of granularities having the same time domain, a granularity G in the set is a bottom granularity with respect to g-rel, if G g-rel H for each granularity H in the set. 
\end{timedom}

**Dicussion:** Given the set of all granularities defined over the time domain (Z; <), and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

@@

## Computation through Calendar Algebra

@@

- **The grouping operation** : Let G be a full-integer labeled granularity, and m a positive integer. The grouping operation $Group_m(G)$ generates a new granularity G, by partitioning the granules of G into m-granule groups and making each group a granule of the resulting granularity.
More precisely, $G = Group_m(G)$ is the full-integer labeled granularity such that for
each integer $i$, $G(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G(j)$.

@@

- **The altering-tick operation** : Let G1, G2 be full-integer labeled granularities, and $l$, $k$, $m$ integers, where G2 partitions G1, and $1 \leq l \leq m$. The altering-tick operation $Alter^{m}_{l,k} (G2, G1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the $l$th granule of each group has $|k|$ additional (or fewer when k < 0) granules of G2.

@@

# Cyclic Time granularities {#sec:circular-gran-def}

Periodicity is very common in all kinds of data. Time granularities which can accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Hence, time granularities can also be **circular** or **aperiodic** depending on if the mapping of the linear granularities is regular or irregular. Examples of circular time granularities include hour of the day, and day of the week, whereas examples for aperiodic time granularities can be day of the month or public holidays.

@@

## Circular {#sec:circular-gran-def}

@@

\begin{equation} \label{eq:eq2}
C_{BG, G}(z) = BG(z\mod k(BG,G)) \forall z \in \mathbb{Z}^+,
BG(z) \neq \phi,
\end{equation}
where,  
z denotes the index set, 
BG denotes the index of the tsibble (bottom granularity), 
G be a linear granularity, periodic with respect to BG,     
k(BG,G) be the period length of the bottom granularity with respect to linear granularity G.  
\end{timedom}

@@

## Quasi-circular {#sec:circular-gran-def}

@@

\begin{equation}\label{eq:eq3}
A_{BG, AG}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert), z \in T_k
\end{equation}
where,  
z denotes the index set (index of BG),
$R'$ denotes the number of granules of AG in each period of BG,  
$T_w$ are the sets of indices of BG describing $AG(w)$ such that $AG(w) = \bigcup_{z \in T_w}BG(z)$, and $T_w = 0$ for $w \epsilon Z_{<0}$  
$\vert T_w \vert$ is the cardinality of set $T_w$.

@@

# Acyclic time granularities {#sec:circular-gran-def}

@@

# Computing cyclic time granularities {#sec:order}

@@

@@

# Data structure

@@

Most commonly used techniques to display a distribution of data include the histogram (Karl Pearson), which shows the prevalence of values grouped into bins and the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. Moreover, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail. “outliers” in letter value plots are those observations beyond the most extreme letter value. The letter values are shown till the depths where the letter values are reliable estimates of their corresponding quantiles and hence might lead to a lot of letter values being shown and leading to overload of information in one plot.

@@

Traditional ways to visualize densities include violin plots [@Hintze1998-zi]. The shape of the violin represents the density estimate of the variable. The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories, enabling easier detection of clusters or bumps within a distribution.

@@

Even when there are no rarely occuring events, number of observations might vary hugely within or across each panel. This might happen due to missing observations in the data or uneven locations of events in time domain. In such cases, the visualizations that use kernel density estimates should be used with caution as sample size would directly affect both the variance of the estimator and the confidence intervals. Measures such as Gini's coefficient may be used to see if sample sizes vary widely across panels.
